[Re] Modeling GABA Alterations in Schizophrenia: A Link Between Impaired Inhibition and Gamma and Beta Auditory Entrainment

### 2017-08-10 article/Metzner-2017.tex
fixed typos, fiyed figure captions, rearranged figures and text

In the original article the authors conduct an MEG study of auditory
entrainment and find changes in gamma and beta range entrainment found
in schizophrenic patients. These changes in oscillatory dynamics in
patients are important for two reasons: first, gamma range oscillations
emerge in a variety of different tasks and are thought to underlie many
cognitive processes (see e.g. \autocite{Fries2005,Fries2015}). Furthermore, impaired gamma
range oscillations in schizophrenic patientas have been found in most of
these tasks and might offer an explanation for the cognitive deficits
found in patients. Therefore, the study of the mechanisms underlying
oscillatory entrainment might shed light on the basic principles that
are responsible for a wide range of deficits in schizophrenic patients.
Second, oscillatory entrainment is a promising candidate for a
neurophysiological biomarker for schizophrenia
\autocite{Siekmeier2015}. In the original article they develop two
models that are able to account for the changes they find in their
experimental data. One is a biophysically detailed network model and the
other one a simplified network model. In both models a prolonged decay
at GABAergic inhibitory synapses causes the reduction in gamma
entrainment and the increase in beta entrainment. The simplified network
model, despite its many simplifications, offers insight into the
mechanisms underlying these entrainment deficits while focusing on a few
key parameters. This makes it easy to distill and understand underlying
dynamics and mechanisms and, at the same time, makes simulations
computationally very inexpensive allowing for an extensive exploration
of the parameter space. Additionally, again because of the simplicity,
the model can easily be extended to study the effect other parameters
(e.g.~sparsity of connections, different populations of inhibitory
neurons, different types of synaptic receptors,\ldots{}), without
loosing much of its simplicity.

We focus on the main results of the original article: an increase in
inhibitory decay time constants leads to a reduction of power in the
gamma range and an increase in power in the beta range, replicating
experimental findings for schizophrenic patients. Therefore, we
reproduce Figures 4,5,6,7,10 and 11 of the original paper. The original
model is implemented using Matlab but the source code is not publicly
available. The model and analysis scripts are implemented using Python
2.7.9. All simulations were run under Ubuntu 15.04.

@@

where \(\alpha _j\) controls excitation and inhibition, i.e. is $+1$ for excitatory synapses and $-1$ for inhibitory ones,
\(g_{jk}\)
is the synaptic strength from cell \(j\) to cell \(k\) and \(s_{jk}\) is
the synaptic gating variable from cell \(j\) to cell \(k\). Synaptic
gating variables are subject to the following dynamics

@@

The model was implemented using Python 2.7.9 using numpy 1.9.3.
Visualisation of results was also done in Python using the matplotlib
module (matplotlib 1.4.3). Furthermore, since the model is
computationally very inexpensive, we did not aim to provide the most
efficient implementation but rather an implementation that is clear, and
easy to understand and use.

@@

\begin{figure}
\centering
\includegraphics{Replication-Figure6.eps}
\caption{Replication of Figure 6: Single trial from the control network.
40 Hz drive. Network entrains to 40 Hz, as can be seen in the frequency
diagram, raster plot and MEG trace.}\label{fig:Vierling6}
\end{figure}

\begin{figure}
\centering
\includegraphics{Replication-Figure7.eps}
\caption{Replication of Figure 7: Single trial from the schizophrenia
network. 40 Hz drive. Network entrains to 40 Hz but also shows a strong
20 Hz component, as can be seen in the frequency diagram, raster plot
and MEG trace. Especially the inhibitory neurons only entrain to a 20 Hz
rhythm (see raster plot).}\label{fig:Vierling7}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{Replication-Figure10.eps}
\caption{Replication of Figure 10: Single trial from the control
network. 20 Hz drive. Network entrains to 20 Hz but also shows a 40 Hz
component, as can be seen in the frequency diagram, raster plot and MEG
trace.}\label{fig:Vierling10}
\end{figure}

\begin{figure}
\centering
\includegraphics{Replication-Figure11.eps}
\caption{Replication of Figure 11: Single trial from the schizophrenia
network. 20 Hz drive. Network entrains to 20 Hz without 40 Hz component,
as can be seen in the frequency diagram, raster plot and MEG trace. note
that the 40 Hz power in the power spectrum is a
harmonic.}\label{fig:Vierling11}
\end{figure}

\subsection{Exploration of the Discrepancies between Original and Reimplementation}
In this section we explore possible scenarios that might explain the less pronounced main features in the reimplementation.

We explore the influence of the background noise. The main mechanism behind the emergence of a 20 Hz component in response
to 40 Hz drive in the schizophrenia network is the prolonged inhibition which suppresses activity in every second 40 Hz cycle. In the absence
of noise the network responds with a pure 20 Hz oscillation (see Figures \ref{fig:NoiseExplorationMEG} left upper panel and \ref{fig:NoiseExplorationPSD} left upper panel). However, the background noise
gives sufficient input to some excitatory cells to overcome this suppression and also fire in between the 20 Hz cycles. This results in a mixed response, 
where power is split between 20 Hz and 40 Hz. The ratio of this split depends on the strength of the noise. Therefore, we asked whether the less pronounced
20 Hz component in our reimplementation might simply come from a too high background noise amplitude. Figures \ref{fig:NoiseExplorationMEG} (left middle and lower panels  and \ref{fig:NoiseExplorationPSD} left middle and lower panels, show the 
response (MEG signal and power spectral density, respectivey) to a 40 Hz drive in the cases, where we scaled down the noise to 75\% and 50\%, respectively. It can clearly be seen that the 20 Hz component increases, when
the noise strength decreases. However, as can be seen in right panels, showing the response of the control network to 20 Hz drive,
is that downscaling of the background noise reduces the 40 Hz component in response to 20 Hz drive. This is not surprising since background noise drives excitatory cells to fire in between 20 Hz cycles in the control network, resulting
in the 40 Hz component. This component is suppressed in the schizophrenia network due to the prolonged inhibition.
Summarising, a difference in the strength of the background noise might be an explanation for the discrepancy between original and reimplementation. However, only a reduction of noise solely for the schizophrenia network 
fully replicates the original results.

\begin{figure}
\centering
\includegraphics{Noise-Exploration-MEG.eps}
\caption{Replication of Figure 11: Single trial from the schizophrenia
network. 20 Hz drive. Network entrains to 20 Hz without 40 Hz component,
as can be seen in the frequency diagram, raster plot and MEG trace. note
that the 40 Hz power in the power spectrum is a
harmonic.}\label{fig:NoiseExplorationMEG}
\end{figure}

\begin{figure}
\centering
\includegraphics{Noise-Exploration-PSD.eps}
\caption{Replication of Figure 11: Single trial from the schizophrenia
network. 20 Hz drive. Network entrains to 20 Hz without 40 Hz component,
as can be seen in the frequency diagram, raster plot and MEG trace. note
that the 40 Hz power in the power spectrum is a
harmonic.}\label{fig:NoiseExplorationPSD}
\end{figure}

@@

@@

Overall, we believe we have faithfully reproduced the main fetaures of
the simple model from \autocite{Vierling2008}. However, we note that
overall features a re a little bit less pronounced in our
reimplementation compared to the original model. We have explored possible explanations for this discrepancy
and conclude that it might come from minor inconsistencies with respect to parameters in the original model. Nevertheless,
the mechanism proposed in the original model could be reproduced and presents a possible explanation for deficits in 
auditory steady-state responses in patients suffering from schizophrenia.

>>>>

In the original article the authors conduct an MEG study of auditory
entrainment and find changes in gamma and beta range entrainment found
in schizophrenic patients. These changes in oscillatory dynamics in
patients are important for two reasons: first, gamma range oscillations
emerge in a variety of different tasks and are thought to underlie many
cognitive processes (see e.g. \autocite{Fries2005,Fries2015}). Furthermore, impaired gamma
range oscillations in schizophrenic patients have been found in most of
these tasks and might offer an explanation for the cognitive deficits
found in patients. Therefore, the study of the mechanisms underlying
oscillatory entrainment might shed light on the basic principles that
are responsible for a wide range of deficits in schizophrenic patients.
Second, oscillatory entrainment is a promising candidate for a
neurophysiological biomarker for schizophrenia
\autocite{Siekmeier2015}. In the original article they develop two
models that are able to account for the changes they find in their
experimental data. One is a biophysically detailed network model and the
other one a simplified network model. In both models a prolonged decay
at GABAergic inhibitory synapses causes the reduction in gamma
entrainment and the increase in beta entrainment. The simplified network
model, despite its many simplifications, offers insight into the
mechanisms underlying these entrainment deficits while focusing on a few
key parameters. This makes it easy to distill and understand underlying
dynamics and mechanisms and, at the same time, makes simulations
computationally very inexpensive allowing for an extensive exploration
of the parameter space. Additionally, again because of the simplicity,
the model can easily be extended to study the effect other parameters
(e.g.~sparsity of connections, different populations of inhibitory
neurons, different types of synaptic receptors,\ldots{}), without
losing much of its simplicity.

We focus on the main results of the original article: an increase in
inhibitory decay time constants leads to a reduction of power in the
gamma range and an increase in power in the beta range, replicating
experimental findings for schizophrenic patients. Therefore, we
reproduce Figures 4, 5, 6, 7, 10 and 11 of the original paper. The original
model is implemented using Matlab but the source code is not publicly
available. The model and analysis scripts are implemented using Python
2.7.9. All simulations were run under Ubuntu 15.04.

@@

where \(n\) is the number of presynaptic neurons, \(\alpha _j\) controls excitation and inhibition, i.e. is $+1$ for excitatory synapses and $-1$ for inhibitory ones,
\(g_{jk}\)
is the synaptic strength from cell \(j\) to cell \(k\) and \(s_{jk}\) is
the synaptic gating variable from cell \(j\) to cell \(k\). Synaptic
gating variables are subject to the following dynamics

@@

The model was implemented using Python 2.7.9 and numpy 1.9.3.
Visualisation of results was also done in Python using the matplotlib
module (matplotlib 1.4.3). Furthermore, since the model is
computationally very inexpensive, we did not aim to provide the most
efficient implementation but rather an implementation that is clear, and
easy to understand and use.

@@

\begin{figure}
\centering
\includegraphics[width=0.9\textwidth]{Replication-Figure6.eps}
\caption{Replication of Figure 6: Single trial from the control network.
40 Hz drive. Network entrains to 40 Hz, as can be seen in the frequency
diagram, raster plot and MEG trace.}\label{fig:Vierling6}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=0.9\textwidth]{Replication-Figure7.eps}
\caption{Replication of Figure 7: Single trial from the schizophrenia
network. 40 Hz drive. Network entrains to 40 Hz but also shows a strong
20 Hz component, as can be seen in the frequency diagram, raster plot
and MEG trace. Especially the inhibitory neurons only entrain to a 20 Hz
rhythm (see raster plot).}\label{fig:Vierling7}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=0.9\textwidth]{Replication-Figure10.eps}
\caption{Replication of Figure 10: Single trial from the control
network. 20 Hz drive. Network entrains to 20 Hz but also shows a 40 Hz
component, as can be seen in the frequency diagram, raster plot and MEG
trace.}\label{fig:Vierling10}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=0.9\textwidth]{Replication-Figure11.eps}
\caption{Replication of Figure 11: Single trial from the schizophrenia
network. 20 Hz drive. Network entrains to 20 Hz without 40 Hz component,
as can be seen in the frequency diagram, raster plot and MEG trace. note
that the 40 Hz power in the power spectrum is a
harmonic.}\label{fig:Vierling11}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=0.65\textwidth]{Noise-Exploration-MEG.eps}
\caption{Raw simulated MEG signals (averaged
over 20 trials) for the control and the schizophrenic network at 20 and 40 Hz driving frequencies 
for three different levels of noise strength (0, 50 and 75 \% of the strength in the standard model).}\label{fig:NoiseExplorationMEG}
\end{figure}

\subsection{Exploration of the Discrepancies between Original and Reimplementation}
In this section we explore possible scenarios that might explain the less pronounced main features in the reimplementation.
We explore the influence of the background noise. The main mechanism behind the emergence of a 20 Hz component in response
to 40 Hz drive in the schizophrenia network is the prolonged inhibition which suppresses activity in every second 40 Hz cycle.

In the absence
of noise the network responds with a pure 20 Hz oscillation (see Figures \ref{fig:NoiseExplorationMEG} left upper panel and \ref{fig:NoiseExplorationPSD} left upper panel). However, the background noise
gives sufficient input to some excitatory cells to overcome this suppression and also fire in between the 20 Hz cycles. This results in a mixed response, 
where power is split between 20 Hz and 40 Hz. The ratio of this split depends on the strength of the noise. Therefore, we asked whether the less pronounced
20 Hz component in our reimplementation might simply come from a too high background noise amplitude. Figures \ref{fig:NoiseExplorationMEG} (left middle and lower panels  and \ref{fig:NoiseExplorationPSD} left middle and lower panels, show the 
response (MEG signal and power spectral density, respectivey) to a 40 Hz drive in the cases, where we scaled down the noise to 75\% and 50\%, respectively. It can clearly be seen that the 20 Hz component increases, when
the noise strength decreases. However, as can be seen in right panels, showing the response of the control network to 20 Hz drive,
is that downscaling of the background noise reduces the 40 Hz component in response to 20 Hz drive. This is not surprising since background noise drives excitatory cells to fire in between 20 Hz cycles in the control network, resulting
in the 40 Hz component. This component is suppressed in the schizophrenia network due to the prolonged inhibition.
Summarising, a difference in the strength of the background noise might be an explanation for the discrepancy between original and reimplementation. However, only a reduction of noise solely for the schizophrenia network 
fully replicates the original results.

\begin{figure}
\centering
\includegraphics[width=\textwidth]{Noise-Exploration-PSD.eps}
\caption{Power spectra of the raw MEG signals from \ref{fig:NoiseExplorationMEG}}\label{fig:NoiseExplorationPSD}
\end{figure}

@@

Overall, we believe we have faithfully reproduced the main fetaures of
the simple model from \autocite{Vierling2008}. However, we note that
overall features are a little bit less pronounced in our
reimplementation compared to the original model. We have explored possible explanations for this discrepancy
and conclude that it might come from minor inconsistencies with respect to parameters in the original model. Nevertheless,
the mechanism proposed in the original model could be reproduced and presents a possible explanation for deficits in 
auditory steady-state responses in patients suffering from schizophrenia.
